How to select predictive models for decision-making or causal inference?

### 2023-12-17 how_to_select_causal_models.tex
iter

Prediction models have long been used in medicine, as with risk
score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
While these have historically been simple models on simple data, this is
changing with progress in machine learning and richer sources of medical data
\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
Complex data is difficult to control and model, but these models are
validated by verifying the accuracy of the prediction on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Crucial to the clinical adoption of a model predicting a health outcome
is that it ``can support decisions about patient care''
\cite{wyatt1995commentary}. Precision medicine is about 
guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? An estimate of the effect of the treatment can be
obtained by contrasting model predictions with and without the treatment,
but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.

@@

Given complex health data, which predictive model is to be most trusted to yield 
valid causal estimates needed to motivate individual treatment decisions?
As no single learner performs best on all data sets, there is a
pressing need for clear guidelines to select outcome models for causal
inference.

>>>>

Prediction models have long been used in medicine, as with risk
score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
While these have historically been simple models on simple data, this is
changing with progress in machine learning and richer medical data
\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
Complex data is difficult to control and model, but these models are
validated by verifying the accuracy of the prediction on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Crucial to the clinical adoption of a model predicting a health outcome
is that it ``can support decisions about patient care''
\cite{wyatt1995commentary}. Precision medicine is about 
guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? An estimate of the effect of the treatment can be
obtained by contrasting model predictions with and without the treatment,
but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.

@@

Given complex health data, which predictive model is to be most trusted to yield 
valid causal estimates needed to motivate individual treatment decisions?
As no single machine-learning method performs best across all data sets, there is a
pressing need for clear guidelines to select outcome models for causal
inference.
